---
title: "Game, Set, Stat: A Comparison of Model Performance Using a Variety of Tennis
  Metrics"
author: "Brady Anderson"
date: "February 25, 2020"
output: 
  beamer_presentation:
    theme: Boadilla
    colortheme: crane
    fig_crop: no
header-includes:
  - \AtBeginDocument{\title[Game, Set, Stat]{Game, Set, Stat - A Comparison of Model Performance Using a Variety of Tennis Metrics}}
  - \usepackage{booktabs}
  - \usepackage{multirow}
  - \usepackage{hyperref}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

#############################################
###### Loading Packages #####################
#############################################
library(fastDummies)
library(lubridate)
require(MASS) 
library(dplyr)
library(leaps)
library(glmnet)
library(chron)
library(ggplot2)
library(glmnetcr)
library(splines)
library(mgcv)
library(caret)
library(leaps)
library(FNN)
library(pls)
library(class)
library(MLmetrics)
library(randomForest)
library(sqldf)
library(knitr)
library(tidyverse)  

#### Load in workspace data ####
load("CC_final.Rdata")
```

## Motivation

- Shift in the men's professional game (ATP)
  - 70% of points between 0 to 4 shots
- Importance of each point
  - World #1 wins $\approx$ 55% of points in a season the past 20 years
  - Wins $90^+$% of matches

## Goals

- Detect variable influence
- Determine model accuracy 
- Measure which model performs best
- Learn several model implementations in R

## Model Performance

- Test Error Rate
- Return on Investment (ROI)

## Dataset - Sources

- Point-by-point shot data
  - [Match Charting Project](http://www.tennisabstract.com/charting/meta.html) (Jeff Sackmann)
- Betting lines data
  - http://www.tennis-data.co.uk/alldata.php

## Dataset - Variables

\begin{table}[h]
\footnotesize
\centering
\begin{tabular}{lcc}
\toprule
Variable & Description & Type\\
\midrule
\textbf{Ace.DF.ratio} & Ratio of Aces to Double Faults & Numeric\\
\textbf{BP.Save.Pct} & Break Points Saved (\%) & Numeric\\
\textbf{Net.Wpct} & Points Won at Net (\%) & Numeric\\
\textbf{Overall.Wpct} & Points Won (\%) & Numeric\\
\textbf{Rank.Pts.Diff} & ATP Ranking Points & Integer\\
\textbf{Serve1.Pct} & 1st Serves In (\%) & Numeric\\
\textbf{Serve1.Wpct} & 1st Serve Points Won (\%) & Numeric\\
\textbf{Serve2.Wpct} & 2nd Serve Points Won (\%) & Numeric\\
\textbf{Serve.Plus1.Pct} & Serves Followed by a Forehand Points Won (\%) & Numeric\\
\textbf{Surface.Clay} & Match Played on Clay Court & Indicator 0-1\\
\textbf{Surface.Grass} & Match Played on Grass Court & Indicator 0-1\\
\textbf{Win.Err.Ratio} & Ratio of Winners to Errors & Numeric\\
\textbf{Zero.to.Four.Wpct} & Points Won between 0-4 Shots (\%) & Numeric\\
\bottomrule
\end{tabular}
\caption{Summary of Variables}
\end{table}
- Use differences in average past data as predictors for test set

## Dataset - Division

- **Training Set (2009-2015)**
  - Build optimal models
  - 565 matches
- **Validation Set (2016-2017)**
  - Tune for optimal hyperparameters ($\lambda$ in Ridge/Lasso)
  - 129 matches
- **Testing Set (2018-2019)**
  - Combine Training/Validation set for final model building
  - Acquire performance metrics
  - 158 matches

## Dataset - Player Distribution

```{r match.chart, fig.height=6}
ggplot(names_final_all,aes(x=Player,y=Matches))+
  geom_bar(stat="identity",fill="blue")+
  labs(title="Charted Match Distribution")+
  xlab("Player")+
  theme(title = element_text(face="bold",color="blue",size=16),axis.text.x = element_text(angle=90,size=8,color="blue",vjust=0.5))+
  scale_x_discrete(limits=players_names,breaks=players_names[seq(1,length(players_names),by=2)])
```

## Methods 

- Logistic Regression
  - Ridge Regression
  - Lasso Regression
  - Principal Components Regression (PCR)
- Linear Discriminant Analysis (LDA)
  - LDA with top PC
- Generalized Additive Model (GAM)
  - GAM with top PC
- Random Forest
- KNN
  - KNN with top PC

## Model Results - Accuracy

\begin{table}[!htbp]
\centering
\begin{tabular}{lll}
\toprule
Model & Type & Error Rate\\
\midrule
Logistic & Full & 44.9\%\\
Logistic & Ridge Regularized & 45.6\%\\
Logistic & Lasso Regularized & 44.9\%\\
Logistic & PCA Compressed & 43.7\%\\
LDA & Full & 46.2\%\\
LDA & PC Compressed & 44.3\%\\
GAM & Full & 45.6\%\\
GAM & PC Compressed & 44.3\%\\
Random Forest & Full & 44.9\%\\
\textbf{KNN} & \textbf{Full} & \textbf{43.0\%}\\
KNN & PC Compressed & 43.7\%\\
\bottomrule
\end{tabular}
\caption{Test Error Rates}
\end{table}

## Model Results - Variable Importance

```{r var.importance}
# How to explain
varImpPlot(rf.train9,main="Variable Importance",col="blue")
# P-values not robust due to correlation issues and non-independence in data used to predict future
  # However - Overall.Wpct, Rank.Pts.Diff only significant ones there 
```


## Betting Strategies

1. All Matches
\begin{equation}
  r_i =
    \begin{cases}
      c, & \text{if $p_i^{model} > 0.5$} \\
      0, & \text{otherwise}
    \end{cases}       
\end{equation}

2. Edge
\begin{equation}
  r_i =
    \begin{cases}
      c, & \text{if $p_i^{model} > p_i^{implied}+0.05$} \\
      0, & \text{otherwise}
    \end{cases}       
\end{equation}

3. Kelly Criterion 
\begin{equation}
  r_i =
    \begin{cases}
      c*\frac{p_i^{model}b_i - (1-p_i^{model})}{b_i}, & \text{if $p_i^{model} > p_i^{implied} + 0.05$} \\
      0, & \text{otherwise}
    \end{cases}       
\end{equation}

- Compare Class 1($<$ 5 matches) vs Class 2 (5$+$ matches)


## ROI Results

\begin{table}[t]
\centering
\fontsize{6}{6}\selectfont
\begin{minipage}{.45\textwidth}
\begin{tabular}{cccc}
\hline
Model                                                                       & Class                 & Strategy & ROI                           \\ \hline
\multicolumn{1}{c|}{{\color[HTML]{009901} }}                                &                       & All      & -3.38                         \\
\multicolumn{1}{c|}{{\color[HTML]{009901} }}                                &                       & Edge     & {\color[HTML]{009901} 5.28}   \\
\multicolumn{1}{c|}{{\color[HTML]{009901} }}                                & \multirow{-3}{*}{(1)} & Kelly    & -5.42                        \\ \cline{2-4} 
\multicolumn{1}{c|}{{\color[HTML]{009901} }}                                &                       & All      & -0.29                         \\
\multicolumn{1}{c|}{{\color[HTML]{009901} }}                                &                       & Edge     & {\color[HTML]{009901} 14.00}  \\
\multicolumn{1}{c|}{\multirow{-6}{*}{{\color[HTML]{009901} Logistic}}}      & \multirow{-3}{*}{(2)} & Kelly    & {\color[HTML]{009901} 4.55}  \\ \hline
\multicolumn{1}{c|}{{\color[HTML]{CB0000} }}                                &                       & All      & -9.08                         \\
\multicolumn{1}{c|}{{\color[HTML]{CB0000} }}                                &                       & Edge     & -9.51                         \\
\multicolumn{1}{c|}{{\color[HTML]{CB0000} }}                                & \multirow{-3}{*}{(1)} & Kelly    & -7.82                        \\ \cline{2-4} 
\multicolumn{1}{c|}{{\color[HTML]{CB0000} }}                                &                       & All      & -8.03                         \\
\multicolumn{1}{c|}{{\color[HTML]{CB0000} }}                                &                       & Edge     & -9.05                         \\
\multicolumn{1}{c|}{\multirow{-6}{*}{{\color[HTML]{CB0000} Ridge}}}         & \multirow{-3}{*}{(2)} & Kelly    & -5.00                        \\ \hline
\multicolumn{1}{c|}{{\color[HTML]{CB0000} }}                                &                       & All      & -11.50                        \\
\multicolumn{1}{c|}{{\color[HTML]{CB0000} }}                                &                       & Edge     & -8.12                         \\
\multicolumn{1}{c|}{{\color[HTML]{CB0000} }}                                & \multirow{-3}{*}{(1)} & Kelly    & -9.34                        \\ \cline{2-4} 
\multicolumn{1}{c|}{{\color[HTML]{CB0000} }}                                &                       & All      & -10.17                        \\
\multicolumn{1}{c|}{{\color[HTML]{CB0000} }}                                &                       & Edge     & -10.28                        \\
\multicolumn{1}{c|}{\multirow{-6}{*}{{\color[HTML]{CB0000} Lasso}}}         & \multirow{-3}{*}{(2)} & Kelly    & -11.27                       \\ \hline
\multicolumn{1}{c|}{}                                                       &                       & All      & -7.36                         \\
\multicolumn{1}{c|}{}                                                       &                       & Edge     & {\color[HTML]{009901} 1.35}   \\
\multicolumn{1}{c|}{}                                                       & \multirow{-3}{*}{(1)} & Kelly    & -13.78                       \\ \cline{2-4} 
\multicolumn{1}{c|}{}                                                       &                       & All      & -1.37                         \\
\multicolumn{1}{c|}{}                                                       &                       & Edge     & {\color[HTML]{009901} 7.92}   \\
\multicolumn{1}{c|}{\multirow{-6}{*}{Logistic PCA}}                         & \multirow{-3}{*}{(2)} & Kelly    & -6.48                         \\ \hline
\multicolumn{1}{c|}{}                                                       &                       & All      & -7.51                         \\
\multicolumn{1}{c|}{}                                                       &                       & Edge     & {\color[HTML]{009901} 1.34}   \\
\multicolumn{1}{c|}{}                                                       & \multirow{-3}{*}{(1)} & Kelly    & -7.82                        \\ \cline{2-4} 
\multicolumn{1}{c|}{}                                                       &                       & All      & -5.77                         \\
\multicolumn{1}{c|}{}                                                       &                       & Edge     & {\color[HTML]{009901} 8.78}   \\
\multicolumn{1}{c|}{\multirow{-6}{*}{LDA Full}}                             & \multirow{-3}{*}{(2)} & Kelly    & {\color[HTML]{009901} 1.03} \\ \hline
\multicolumn{1}{c|}{{\color[HTML]{CB0000} }}                                &                       & All      & -9.16                         \\
\multicolumn{1}{c|}{{\color[HTML]{CB0000} }}                                &                       & Edge     & -6.54                         \\
\multicolumn{1}{c|}{{\color[HTML]{CB0000} }}                                & \multirow{-3}{*}{(1)} & Kelly    & -16.29                       \\ \cline{2-4} 
\multicolumn{1}{c|}{{\color[HTML]{CB0000} }}                                &                       & All      & -3.76                         \\
\multicolumn{1}{c|}{{\color[HTML]{CB0000} }}                                &                       & Edge     & -1.88                         \\
\multicolumn{1}{c|}{\multirow{-6}{*}{{\color[HTML]{CB0000} LDA PCA}}}       & \multirow{-3}{*}{(2)} & Kelly    & -10.22   \\
\hline
\end{tabular}
\end{minipage}
\begin{minipage}{.45\textwidth}
\begin{tabular}{cccc}
\hline
Model                                                                       & Class                 & Strategy & ROI 
                        \\ \hline
\multicolumn{1}{c|}{{\color[HTML]{009901} }}                                &                       & All      & -3.00                         \\
\multicolumn{1}{c|}{{\color[HTML]{009901} }}                                &                       & Edge     & {\color[HTML]{009901} 4.01}   \\
\multicolumn{1}{c|}{{\color[HTML]{009901} }}                                & \multirow{-3}{*}{(1)} & Kelly    & -5.01                        \\ \cline{2-4} 
\multicolumn{1}{c|}{{\color[HTML]{009901} }}                                &                       & All      & {\color[HTML]{009901} 0.22}   \\
\multicolumn{1}{c|}{{\color[HTML]{009901} }}                                &                       & Edge     & {\color[HTML]{009901} 10.64}  \\
\multicolumn{1}{c|}{\multirow{-6}{*}{{\color[HTML]{009901} GAM Full}}}      & \multirow{-3}{*}{(2)} & Kelly    & -0.67                         \\ \hline
\multicolumn{1}{c|}{}                                                       &                       & All      & -9.13                         \\
\multicolumn{1}{c|}{}                                                       &                       & Edge     & {\color[HTML]{009901} 0.93}   \\
\multicolumn{1}{c|}{}                                                       & \multirow{-3}{*}{(1)} & Kelly    & -13.38                         \\ \cline{2-4} 
\multicolumn{1}{c|}{}                                                       &                       & All      & -5.04                         \\
\multicolumn{1}{c|}{}                                                       &                       & Edge     & {\color[HTML]{009901} 6.21}   \\
\multicolumn{1}{c|}{\multirow{-6}{*}{GAM PCA}}                              & \multirow{-3}{*}{(2)} & Kelly    & -9.01                         \\ \hline
\multicolumn{1}{c|}{{\color[HTML]{009901} }}                                &                       & All      & -2.41                         \\
\multicolumn{1}{c|}{{\color[HTML]{009901} }}                                &                       & Edge     & {\color[HTML]{009901} 4.21}   \\
\multicolumn{1}{c|}{{\color[HTML]{009901} }}                                & \multirow{-3}{*}{(1)} & Kelly    & -4.87                         \\ \cline{2-4} 
\multicolumn{1}{c|}{{\color[HTML]{009901} }}                                &                       & All      & -0.81                         \\
\multicolumn{1}{c|}{{\color[HTML]{009901} }}                                &                       & Edge     & {\color[HTML]{009901} 10.26}  \\
\multicolumn{1}{c|}{\multirow{-6}{*}{{\color[HTML]{009901} Random Forest}}} & \multirow{-3}{*}{(2)} & Kelly    & {\color[HTML]{009901} 0.21}  \\ \hline
\multicolumn{1}{c|}{}                                                       &                       & All      & -7.49                         \\
\multicolumn{1}{c|}{}                                                       &                       & Edge     & -1.80                         \\
\multicolumn{1}{c|}{}                                                       & \multirow{-3}{*}{(1)} & Kelly    & -7.44                        \\ \cline{2-4} 
\multicolumn{1}{c|}{}                                                       &                       & All      & -7.83                         \\
\multicolumn{1}{c|}{}                                                       &                       & Edge     & {\color[HTML]{009901} 1.41}   \\
\multicolumn{1}{c|}{\multirow{-6}{*}{KNN Full}}                             & \multirow{-3}{*}{(2)} & Kelly    & -1.40                        \\ \hline
\multicolumn{1}{c|}{}                                                       &                       & All      & -12.50                        \\
\multicolumn{1}{c|}{}                                                       &                       & Edge     & {\color[HTML]{009901} 0.74}   \\
\multicolumn{1}{c|}{}                                                       & \multirow{-3}{*}{(1)} & Kelly    & -0.81                        \\ \cline{2-4} 
\multicolumn{1}{c|}{}                                                       &                       & All      & -10.39                        \\
\multicolumn{1}{c|}{}                                                       &                       & Edge     & {\color[HTML]{009901} 9.29}   \\
\multicolumn{1}{c|}{\multirow{-6}{*}{KNN PCA}}                              & \multirow{-3}{*}{(2)} & Kelly    & {\color[HTML]{333333} -4.89}  \\
\hline
\end{tabular}
\end{minipage}
\end{table}

## Conclusion

- KNN was the most accurate model
- Logistic Regression and Random Forest produced the highest ROI
- Overall $\Rightarrow$ Logistic Regresion or Random Forest

## Limitations

- Non-representative data
  - More matches from top players
- Limited data for some players
  - Irregularity in time for averaging
- Charting differences

## Further Study

- Complete data
- Model adjustments
  - Time series incorporation
- WTA models

## Work Cited

\fontsize{10pt}{6.2}\selectfont

1. James, G., Witten, D., Trevor, H., & Tibshirani, R. (2013). *An Introduction to Statistical Learning*. New York : Springer.

2. J. Kelly. "A new interpretation of information rate". *IRE Transactions on Information Theory*, 2(3):917–926, 1956.

3. Lopez, Michael J., and Gregory J. Matthews. "Building an NCAA men’s basketball predictive model and quantifying its success." *Journal of Quantitative Analysis in Sports* 11.1 (2015): 5-12.

4. O’Shannessy, Craig. “The First 4 Shots.” (2017).

5. Sackmann, Jeff. “Tennis Abstract: Match Charting Project Metadata.” \url{tennisabstract.com}, (2019), \url{www.tennisabstract.com/charting/meta.html}.

6. Sipko, Michal, and William Knottenbelt. "Machine learning for the prediction of professional tennis matches." *MEng computing-final year project, Imperial College London* (2015).

7. Tennis Betting | Tennis Results | Tennis Odds. \url{http://www.tennis-data.co.uk/alldata.php} (2019). 

## Thank you

- Questions?
